Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-29 main.tex
Update on Overleaf.

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
\HIDE{From FORTRAN to Numpy, arrays have revolutionized how we express computation.  Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.  However, arrays can only handle dense rectilinear integer grids. Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry. We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression. }

Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays. Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy. Furthermore, multi-dimensional arrays are the highest-performing data structure with a long history of investment and innovation, from hardware support to compiler
technology.  Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
However, a large portion of the data, either obtained from nature or generated by humans, have structure.

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes. So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data. Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

In this paper, we describe a compiler, Finch,
which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques in areas such as
sparse tensors, databases, and lossless compression.
We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  
\end{abstract}

@@

\section Frequently Requested Finch kernels

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\input{sections/abstract}

@@

\section{Frequently Requested Finch kernels}

### 2024-03-29 sections/background.tex
formatting
Update on Overleaf.

\section{Background}
\subsection{Looplets}
Finch represents iteration patterns using Looplets, a language that decomposes
datastructure iterators hierarchically. Looplets represent the control-flow
structures needed to iterate over any given datastructure, or multiple
datastructures simultaneously. In particular, looplets are good at lifting code
to the highest loop level that it's needed and subdividing iteration
hierarchically in coordinate space. Because looplets are compiled with
progressive lowering, structure-specific mathematical optimizations such as
integrals, multiply by zero, etc. can be implemented using simple compiler
passes like term rewriting and constant propagation during the intermediate
lowering stages. \cite{ahrens_looplets_2023}

@@

\begin{figure}[ht]
    \begin{minipage}[c]{0.6\linewidth}
        \paragraph{$\finchlookup(seek, body)$} The Lookup Looplet represents a
        randomly accessible region of an iterator. The body of the lookup is
        understood to have one less dimension than the lookup itself, as we have
        already ``looked up'' that index in the tensor by the time we reach the
        body. \texttt{seek(i)} is a function that updates state to the given
        index.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-lookup.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        \paragraph{$\finchrun(body)$} The Run Looplet represents a constant
        region of an iterator. The body of the run is understood to have one
        less dimension than the lookup itself, as all of the bodies are
        identical.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-run.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        \paragraph{$\finchphase(c:d, body)$} The Phase Looplet represents a
        restriction of the range on which a loop should execute, and allows us
        to succinctly express the ranges on which children of compound looplets
        are defined.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-phase.png}
    \end{minipage}
    \begin{minipage}[c]{0.6\linewidth}
        \paragraph{$\finchswitch(cond, head, tail)$} The Switch Looplet allows
        us to specialize the body of a looplet based on a condition, evaluated
        in the embedding context. If the condition is true, we use `head`,
        otherwise we use `tail`. Switch has a high lowering priority so we can
        see what's inside of it and lower that appropriately. This also lifts
        the condition as high as possible into the loop nest. The condition is
        assumed to evaluate to a boolean.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-switch.png}
    \end{minipage}
    \begin{minipage}[c]{0.6\linewidth}
        \paragraph{$\finchthunk(preamble, body, epilogue)$} The Thunk Looplet
        allows us to cache certain computations in the state under which the
        body will execute. This is useful for computing and caching the results
        of expensive computations.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-thunk.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        \paragraph{$\finchsequence(head, tail)$} The Sequence looplet represents the
        concatenation of two looplets. Both arguments must be phase looplets, and
        are assumed to be nonoverlapping, covering, and in order.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-sequence.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        \paragraph{$\finchspike(body, tail)$} The Spike Looplet represents a run
        followed by a single value. In this paper, Spike will be considered a
        shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)),
        \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are
        handled with special care, since they are an opportunity to align the
        final run to the end of the root loop extent, without using any special
        bounds inference.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-spike.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        \paragraph{$\finchstepper([seek], next, body)$} The stepper looplet
        represents a variable number of looplets, concatenated. Since our
        looplets may be skipped over due to conditions or various rewrites, the
        $seek$ function allows us to fast-forward the state to the start of the
        root loop extent when it comes time to lower the stepper. The $next$
        function advances the state to the next iteration of the stepper. 
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-stepper.png}
    \end{minipage}
    \caption{The looplet language, as understood in a correct execution of a Finch program.}
\end{figure}

\subsection{FiberTrees}

\subsection{Concordant Iteration}

\subsection{Protocols}

>>>>

\section{Background On Looplets}
Finch represents iteration patterns using Looplets, a language that decomposes
datastructure iterators hierarchically. Looplets represent the control-flow
structures needed to iterate over any given datastructure, or multiple
datastructures simultaneously. In particular, looplets are good at lifting code
to the highest loop level that it's needed and subdividing iteration
hierarchically in coordinate space. Because looplets are compiled with
progressive lowering, structure-specific mathematical optimizations such as
integrals, multiply by zero, etc. can be implemented using simple compiler
passes like term rewriting and constant propagation during the intermediate
lowering stages. \cite{ahrens_looplets_2023}

@@

\begin{figure}[ht]
    \footnotesize
    \begin{minipage}[c]{0.6\linewidth}
        $\finchlookup(seek, body)$: The Lookup Looplet represents a
        randomly accessible region of an iterator. The body of the lookup is
        understood to have one less dimension than the lookup itself, as we have
        already ``looked up'' that index in the tensor by the time we reach the
        body. \texttt{seek(i)} is a function that updates state to the given
        index.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-lookup.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        $\finchrun(body)$: The Run Looplet represents a constant
        region of an iterator. The body of the run is understood to have one
        less dimension than the lookup itself, as all of the bodies are
        identical.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-run.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        $\finchphase(c:d, body)$: The Phase Looplet represents a
        restriction of the range on which a loop should execute, and allows us
        to succinctly express the ranges on which children of compound looplets
        are defined.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-phase.png}
    \end{minipage}
    \begin{minipage}[c]{0.6\linewidth}
        $\finchswitch(cond, head, tail)$: The Switch Looplet allows
        us to specialize the body of a looplet based on a condition, evaluated
        in the embedding context. If the condition is true, we use `head`,
        otherwise we use `tail`. Switch has a high lowering priority so we can
        see what's inside of it and lower that appropriately. This also lifts
        the condition as high as possible into the loop nest. The condition is
        assumed to evaluate to a boolean.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-switch.png}
    \end{minipage}
    \begin{minipage}[c]{0.6\linewidth}
        $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet
        allows us to cache certain computations in the state under which the
        body will execute. This is useful for computing and caching the results
        of expensive computations.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-thunk.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        $\finchsequence(head, tail)$: The Sequence looplet represents the
        concatenation of two looplets. Both arguments must be phase looplets, and
        are assumed to be nonoverlapping, covering, and in order.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-sequence.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        $\finchspike(body, tail)$ The Spike Looplet represents a run
        followed by a single value. In this paper, Spike will be considered a
        shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)),
        \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are
        handled with special care, since they are an opportunity to align the
        final run to the end of the root loop extent, without using any special
        bounds inference.
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-spike.png}
    \end{minipage}

    \begin{minipage}[c]{0.6\linewidth}
        $\finchstepper([seek], next, body)$ The stepper looplet
        represents a variable number of looplets, concatenated. Since our
        looplets may be skipped over due to conditions or various rewrites, the
        $seek$ function allows us to fast-forward the state to the start of the
        root loop extent when it comes time to lower the stepper. The $next$
        function advances the state to the next iteration of the stepper. 
    \end{minipage}%
    \begin{minipage}[c]{0.4\linewidth}
        \centering
        \includegraphics[scale=0.25]{Looplets-stepper.png}
    \end{minipage}
    \caption{The looplet language, as understood in a correct execution of a Finch program.}
\end{figure}

### 2024-03-29 sections/intro.tex
Update on Overleaf.

\begin{table}[h!]
\noindent
\flushleft
\begin{minipage}[t]{.49\textwidth}
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{StructTensor} & \rothead{Finch} \\
  \hline
  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  % High-Level API           &            & \checkmark &            &            &            & \checkmark \\
  % Automatic API Fusion     &            &            &            &            &            & \checkmark \\
%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            &           \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
    Arbitrary Read/Write              &    \checkmark        &  &            &  &            &\checkmark \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
\end{minipage} 
\begin{minipage}[t]{.49\textwidth}
  \flushright
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{StructTensor} & \rothead{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse               &            & \checkmark &            & \checkmark &            &\checkmark \\
  Sparse                   &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{minipage}
\end{table}

@@

\begin{enumerate}
\item More complex array structures than ever before. We are the first to extend level-by-level hierarchical descriptions to capture banded, triangular, run-length-encoded, or sparse datasets, and any combination thereof.
%
We have chosen a set of level formats that completely captures all combinations of relevant structural properties (zeros, repeated values, and/or blocks).
%
Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level. 
%
\item A rich structured array programming language with for-loops and complex control flow constructs at the same level of productivity of dense arrays. 
%
To our knowledge, the Finch programming language is the first to support if-conditions, early breaks, and multiple left hand sides over structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands.
%
\item A compiler that specializes programs to data structures automatically, facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures. Finch reliably utilizes the key properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
%
\item Our compiler is highly extensible, evidenced by the variety of level formats and control flow constructs we implement in this work.
%
For example, Finch has been extended to support real-valued array indices with continuous arrays \cite{won2024continuous}.
%
\item We evaluate the efficiency, flexibility, and expressability of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>
\end{enumerate}

>>>>

\begin{table}[h!]
\noindent
\flushleft
\begin{minipage}[t]{.49\textwidth}
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Einsums/Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  % High-Level API           &            & \checkmark &            &            &            & \checkmark \\
  % Automatic API Fusion     &            &            &            &            &            & \checkmark \\
%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            &           \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Unrestricted Read/Write              &    \checkmark        &  &            &  &            &  \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
\end{minipage} 
\begin{minipage}[t]{.49\textwidth}
  \flushright
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse Operand              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Multiple Sparse  Operands                 &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{minipage}
\end{table}

@@

\begin{enumerate}
\item More complex array structures than ever before. We are the first to extend level-by-level hierarchical descriptions to capture banded, triangular, run-length-encoded, or sparse datasets, and any combination thereof.
%
We have chosen a set of level formats that completely captures all combinations of relevant structural properties (zeros, repeated values, and/or blocks).
%
Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level. 
%
\item A rich structured array programming language with for-loops and complex control flow constructs at the same level of productivity of dense arrays. 
%
To our knowledge, the Finch programming language is the first to support if-conditions, early breaks, and multiple left hand sides over structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands.
%
\item A compiler that specializes programs to data structures automatically, facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures. Finch reliably utilizes four key properties of structure, such as structural zeros, repeated values, clustered non-zeros, and singletons.
%
\item Our compiler is highly extensible, evidenced by the variety of level formats and control flow constructs we implement in this work.
%
For example, Finch has been extended to support real-valued array indices with continuous arrays \cite{won2024continuous}.
%
\item We evaluate the efficiency, flexibility, and expressability of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>

\end{enumerate}

This advances the state of the art over the Looplets work, as the Looplets presented a way to merge iterators over multiple single dimensional structures, whereas we present a way to combine single-dimensional looplet structures into multi-dimensional tensors and operate on them with a programming language with fully-featured control flow.
